Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2022-12-22 paper-1/manuscript.tex
Rev Fig2-3, minus Fig2 text

\begin{figure*}%
    \centering
    \includegraphics{fig2.pdf}
    \caption{}
    \label{fig2}
\end{figure*}
\textbf{Fig. 2} Deviance detection in vitro and in silico. \textbf{a} Response of a 2 $\times$ 4~mm neuronal cell culture on a multi-electrode array, adapted from \cite{Kubota2021-dx}. Curves represent the average number of spikes per electrode in each 10~ms bin in response to target stimulation in the indicated sequence. Unlike in our study, oddball sequences here were split 10\%/90\%, and 10 equally frequent stimuli were used for the msc sequence. \textbf{b} Sample model response, showing average network-wide spike counts after stimulation with target (A) in the indicated sequence (same color code as panel \textbf{a}). Both in vitro and in the model, the dev response is very large, followed by an intermediate msc response and a low std response. \textbf{c} Mismatch, SSA and deviance detection indices (see Equation \ref{eqn-ddi}) across networks and stimuli (n = 120). In this and later boxplots, the median is indicated with an orange line, notches indicate the 95\% confidence interval for the median calculated by bootstrap with 10000 iterations, boxes indicate the inter-quartile range (IQR), whiskers extend to 1.5 times the IQR or to the data extrema, whichever is less, and fliers indicate individual data points beyond the whisker end points

@@

Our first aim was to replicate deviance detection properties reported in dissociated neuronal cell culture by \cite{Kubota2021-dx}, reproduced in Figure \ref{fig2}a. In their case, cell cultures -- like our model, randomly connected networks with two-dimensional extent -- were stimulated at 10 different sites in msc sequences, and at two sites with 10\%/90\% probability in dev and std sequences. Deviance detection was assessed analogously to our definition of DDI and reached an average level of around 0.3 across two cultures.

In our model, we could readily identify networks and stimuli whose responses developed in a qualitatively similar fashion, with high activity in response to target in the dev sequence, low activity in the std sequence, and intermediate activity in the msc sequence, as illustrated with an example in Figure \ref{fig2}b. The model responses developed on a compressed time scale compared to neuronal culture, which we tentatively attribute to unfaithful aspects of the model, including a lack of conduction delay or spontaneous activity, a much smaller number of participating neurons, and structural differences.

Across networks and stimuli (Figure \ref{fig2}c), we found that our model reliably produced positive mismatch responses (t = 11.3, p = 6.7e-21; one-sided t-test, n = 120) while both SSA index (t = 7.84, p = 1.05e-12) and DDI (t = 6.49, p = 1.05e-09) were also positive. The average DDI value reached around 0.1, which was less than the values reported by \cite{Kubota2021-dx}, but note that our target stimuli were presented more frequently (20\%) than theirs (10\%) in the dev and msc sequences.

\begin{figure*}%
    \centering
    \includegraphics{fig3.pdf}
    \caption{}
    \label{fig3}
\end{figure*}
\textbf{Fig. 3} Deviance detection under model ablation. \textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. The color code is the same as in Figure \ref{fig2}. \textbf{b} SSA indices across all networks and stimuli. Asterisks indicate data greater than 0, established by t-test with significance level 0.05, see main text for detailed statistics. \textbf{c} Deviance detection indices across all networks and stimuli.

@@

The SSA index (Figure \ref{fig3}b) was exactly 0 as expected in the ablation with no plasticity, and positive with any plasticity applied (STD only: t = 3.17, p = 0.000983, median 0.0039; TA only: t = 6.74, p = 3e-10, median 0.097; STD+TA: t = 7.84, p = 1.05e-12, median 0.135; one-sided t-tests, n = 120). Since the SSA index measures the degree of adaptation to frequent stimulus presentation, this indicates that both plasticity mechanisms were capable of imbuing the networks with some level of history dependence.

Finally, across networks and stimuli, the DDI (Figure \ref{fig3}c) was exactly 0 with no plasticity, not significantly different from 0 with STD only (t = -1.89, p = 0.0609, median -2.1e-04; two-sided t-test, n = 120), but clearly positive for TA only (t = 5.52, p = 1.01e-07, median 0.077; one-sided t-test, n = 120) and the full model (t = 6.49, p = 1.05e-09, median 0.114). To our surprise, the DDI of the full model was significantly greater than that of either plasticity mechanism alone (STD: t = 6.7, p = 4.53e-10; TA: t = 3.2, p = 0.00103), and greater too than the sum of the DDI values across both STD only and TA only models (t = 3.6, p = 0.000199). This suggests that TA and STD act not as independent mechanisms in this paradigm, but rather interact constructively to enhance the network's ability to encode input regularities.

>>>>

\begin{figure*}%
    \centering
    \includegraphics{fig2.pdf}
    \caption{}
    \label{fig2}
\end{figure*}
\textbf{Fig. 2} Deviance detection in our model. \textbf{a} Sample network target response intensity by trial type. Each horizontal line of the plot summarizes the network response in one target trial in terms of spikes per integration time step. Target trials are taken from the std (400), msc (100), and dev (100) sequences, and sorted by sequence time (earliest trials are at the bottom of each section). \textbf{b} Response summary, showing average network-wide spike counts in target trials in each sequence. The underlying data are the same as in panel a. Note how the dev response is very large, followed by an intermediate msc response and a low std response. \textbf{c} Mismatch, adaptation, and deviance detection indices (see Equation \ref{eqn-ddi}) across networks and stimuli (n = 120). In this and later boxplots, the median is indicated with an orange line, notches indicate the 95\% confidence interval for the median calculated by bootstrap with 10000 iterations, boxes indicate the inter-quartile range (IQR), whiskers extend to 1.5 times the IQR or to the data extrema, whichever is less, and fliers indicate individual data points beyond the whisker end points

@@

% Our first aim was to replicate deviance detection properties reported in dissociated neuronal cell culture by \cite{Kubota2021-dx}, reproduced in Figure \ref{fig2}a. In their case, cell cultures -- like our model, randomly connected networks with two-dimensional extent -- were stimulated at 10 different sites in msc sequences, and at two sites with 10\%/90\% probability in dev and std sequences. Deviance detection was assessed analogously to our definition of DDI and reached an average level of around 0.3 across two cultures.

% In our model, we could readily identify networks and stimuli whose responses developed in a qualitatively similar fashion, with high activity in response to target in the dev sequence, low activity in the std sequence, and intermediate activity in the msc sequence, as illustrated with an example in Figure \ref{fig2}b. The model responses developed on a compressed time scale compared to neuronal culture, which we tentatively attribute to unfaithful aspects of the model, including a lack of conduction delay or spontaneous activity, a much smaller number of participating neurons, and structural differences.

% Across networks and stimuli (Figure \ref{fig2}c), we found that our model reliably produced positive mismatch responses (T = 6.67e+03, p = 9.26e-16; one-sided Wilcoxon signed-rank test, n = 120) while both SSA index (T = 6.19e+03, p = 9.76e-12) and DDI (T = 5.97e+03, p = 4.75e-10) were also positive. The average DDI value reached around 0.1, which was less than the values reported by \cite{Kubota2021-dx}, but note that our target stimuli were presented more frequently (20\%) than theirs (10\%) in the dev and msc sequences.

\begin{figure*}%
    \centering
    \includegraphics{fig3.pdf}
    \caption{}
    \label{fig3}
\end{figure*}
\textbf{Fig. 3} Deviance detection under model ablation. \textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. Lines show the average (trial mean +- SEM, left axis) network-wide spike counts in target trials in each sequence. Boxes summarize the spike count totals per trial (right axis), the means of which form the basis for index calculations (see panel c). Note the different axis scaling on the top row (without TA) and the bottom row (with TA). \textbf{b} Adaptation indices $<$msc - std$>$ across all networks and stimuli. Asterisks indicate data greater than 0, established with a Wilcoxon signed-rank test at a significance level 0.05. See main text for detailed statistics. \textbf{c} Deviance detection indices $<$dev - msc$>$ across all networks and stimuli

@@

The SSA index (Figure \ref{fig3}b) was exactly 0 as expected in the ablation with no plasticity, and positive with any plasticity applied (STD only: T = 4.97e+03, p = 0.000222, median 0.0039; TA only: T = 5.99e+03, p = 3.3e-10, median 0.097; STD+TA: T = 6.19e+03, p = 9.76e-12, median 0.135; one-sided Wilcoxon signed-rank tests, n = 120). Since the SSA index measures the degree of adaptation to frequent stimulus presentation, this indicates that both plasticity mechanisms were capable of imbuing the networks with some level of history dependence.

Finally, across networks and stimuli, the DDI (Figure \ref{fig3}c) was exactly 0 with no plasticity, not significantly different from 0 with STD only (T = 3.2e+03, p = 0.262, median -2.1e-04; two-sided Wilcoxon, n = 120), but clearly positive for TA only (T = 5.85e+03, p = 3.1e-09, median 0.077; one-sided Wilcoxon, n = 120) and the full model (T = 5.97e+03, p = 4.75e-10, median 0.114). To our surprise, the DDI of the full model was significantly greater than that of either plasticity mechanism alone (STD: T = 6.1e+03, p = 4.95e-11; TA: T = 4.87e+03, p = 0.000604), and greater too than the sum of the DDI values across both STD only and TA only models (T = 5.12e+03, p = 4.52e-05). This suggests that TA and STD act not as independent mechanisms in this paradigm, but rather interact constructively to enhance the network's ability to encode input regularities.
